Supplemental data to ``\emph{k}-mer-based GWAS enhances causal variant and candidate gene discovery in soybean''

### 2023-02-28 additional_files/additional_file_1.tex
Adjust the contents of the main and supplementary results to reduce redundancy ; figure order has yet to be adjusted accordingly

% Detailed results that are not meant to fit the main paper
\section*{Detailed results}
\label{annexe-sv-gwas-detailed-results}

This section provides a more detailed per-trait account of some of the results obtained
by the GWAS analyses. Aspects of the results that were
already discussed in the main text are not repeated here; in such cases,
readers are simply redirected to specific tables and figures containing
detailed results from all four GWAS approaches. A summary of the signals found
at all loci using all four methods is presented in Table \ref{signals_table}.

\subsection*{Flower color}
\label{sv-gwas-flower-color}

We conducted a GWAS analysis contrasting purple and white flower colors
(Table \ref{FLOWER.COLOR_gwas_table}). Of the two classical loci reported by
\cite{bandillo2017}, \textit{W1} was detected by all four GWAS approaches whereas
\textit{L1} was detected by none of the approaches (Figure
\ref{flower_color_manhattan}). These results are unsurprising as W1 is the
major contributor to flower color in soybean, whereas the \textit{L1} locus
(typically associated with pod color) was detected with a much less significant
\textit{p}-value (see Table \ref{signals_table}) and only conjectured as
associated with flower color by \cite{bandillo2017}. Detailed results for the
\textit{W1} locus can be viewed in figures \ref{flower-color-main-figure},
\ref{flower_color_W1_signal} and \ref{flower_color_W1_gene}.

The \textit{k}-mer GWAS analysis also identified presumably spurious significant
\textit{k}-mers scattered throughout the genome (Figure
\ref{flower_color_manhattan}). An analysis of the pairwise LD between a subset
of those \textit{k}-mers showed that most \textit{k}-mers formed a single LD block
linked to the \textit{W1} locus (Figure \ref{flower_color_ld}). A small LD block
occurred on chromosome Gm09, however it was in moderate LD with the one on chromosome
Gm13 and therefore probably does not represent a separate locus.

\subsection*{Pubescence color}
\label{sv-gwas-pubescence-color}

We conducted two different GWAS analyses on pubescence color. The first analysis
contrasted gray, tawny and light tawny pubescence colors (Table
\ref{PUBESCENCE.COLOR_gwas_table}) and detected a signal overlapping the
\emph{T} locus using all four approaches (Figure
\ref{pubescence_color_all_manhattan}). Detailed results at the \textit{T} locus
can be viewed in figures \ref{pubescence_color_all_T_signal},
\ref{pubescence_color_all_T_gene} and \ref{pubescence_color_all_T_kmers}. We
observed that not all accessions with gray pubescence harbored the deleted
nucleotide, which could indicate either that other causal variants are resulting in
a non-functional F3’H in this population, or that the causal
variant lies elsewhere.

This first GWAS analysis also revealed a second locus on chromosome Gm16
(Figure \ref{pubescence_color_all_manhattan}). Upon closer analysis, this locus
appeared to be in strong LD with \emph{k}-mers mapping to chromosome Gm01 and
most importantly in moderate LD with the \emph{T} locus (Figure
\ref{pubescence_color_all_ld}), which suggests that this does not represent an
independent locus controlling pubescence color. This interpretation is also
supported by the fact that the signal at that locus only spans about 10 kb on
chromosome Gm16.

In the presence of a functional allele at the \emph{T} locus, the genotype at
the \emph{Td} locus influences the exact shade of pubescence color.  Therefore,
we conducted a second GWAS analysis contrasting accessions with tawny and light
tawny pubescence in order to ignore the epistatic effect of the \emph{T} locus
and instead focus on the effect of the \emph{Td} locus on pubescence color
(Table \ref{PUBESCENCE.COLOR_gwas_table}). Detailed results at this locus
can be viewed in figures \ref{pubescence-color-main-figure}, \ref{pubescence_color_nogray_Td_signal} and
\ref{pubescence_color_nogray_Td_gene}).

\subsection*{Seed coat color}
\label{sv-gwas-seed-coat-color}

We conducted two different GWAS analyses on seed coat color. The first GWAS
included five different phenotypes observed in the dataset (Table
\ref{SEED.COAT.COLOR_gwas_table}) and resulted in the detection of strong
signals overlapping the \emph{I} locus by all four approaches (Figure
\ref{seed_coat_color_all_manhattan}). All approaches also detected
signals at the \emph{G} locus. In addition, the \emph{k}-mer approach also
detected weak signals overlapping the \emph{T} and \emph{O} loci. However, the
\emph{T} and \emph{O} loci are likely chance associations given the low \textit{p}-value
associated with the \emph{T} locus and the proximity of the \emph{O} locus to
the \emph{I} locus. Detailed results of this first GWAS and the \textit{I}
locus can be viewed in figures \ref{seed_coat_color_all_manhattan} and
\ref{seed_coat_color_all_I_signal}.

The second GWAS analysis on seed coat color focused on the subset of accessions
with yellow or green seed coat only (Table \ref{SEED.COAT.COLOR_gwas_table}),
as this difference is controlled by the \emph{G} locus. Detailed results of
this analysis can be viewed in figures \ref{seed-coat-color-main-figure},
\ref{seed_coat_color_greenyellow_manhattan},
\ref{seed_coat_color_greenyellow_G_signal} and
\ref{seed_coat_color_greenyellow_G_gene}.

\subsection*{Stem termination type}
\label{sv-gwas-stem-termination}

We conducted two different GWAS analyses on stem termination type. The first GWAS
included all phenotypes observed in the dataset (determinate, semi-determinate,
and indeterminate; Table \ref{STEM.TERMINATION.TYPE_gwas_table}). The results
of this analysis at the \textit{Dt1} locus can be viewed in figures \ref{stem_termination_all_manhattan},
\ref{stem_termination_all_Dt1_signal}, \ref{stem_termination_all_Dt1_gene} and
\ref{stem_termination_all_Dt1_kmers}.

The second GWAS analysis included only
semi-determinate and indeterminate phenotypes. We performed this analysis in an attempt to also detect signals
at the \emph{Dt2} locus (Table \ref{STEM.TERMINATION.TYPE_gwas_table}).
However, this was unsuccessful using all four GWAS
approaches (Figure \ref{stem_termination_sn_manhattan}). Similarly, we failed to
detect signals at the \textit{E3} locus, which is more commonly associated
with maturity but was mentioned as associated with stem termination type by
\cite{bandillo2017}.

\subsection*{Hilum color}
\label{sv-gwas-hilum-color}

We conducted three different GWAS analyses on hilum color. The first GWAS
included seven colors (Table \ref{HILUM.COLOR_gwas_table}) and detected signals
at the \emph{I}, \emph{T} and \emph{R} loci using all four approaches
(Figure \ref{hilum_color_all_manhattan}). No signals
were detected at the \emph{W1} and \emph{O} loci studied by
\cite{bandillo2017}. The detailed results for the \textit{I} locus can be viewed
in figure \ref{hilum_color_all_I_signal} whereas those for the \textit{T} locus
can be viewed in figures \ref{hilum_color_all_T_signal}, \ref{hilum_color_all_T_gene}
and \ref{hilum_color_all_T_kmers}.

Although \cite{bandillo2017} listed the \textit{W1} locus (typically associated
with flower color) as putatively associated with hilum color, we found no
evidence of association in our results.

The second GWAS analysis contrasting accessions with either black or brown hilum
color was better suited at resolving the signal at the \emph{R} locus (Table
\ref{HILUM.COLOR_gwas_table}). Accordingly, this GWAS analysis detected strong
signals overlapping the \emph{R} locus and no other classical loci using all
four approaches (Figure \ref{hilum_color_blackbrown_manhattan}).
\cite{gillman2011} identified the R2R3 MYB transcription factor Glyma.09g235100
as the molecular basis for the \emph{R} locus and documented four different
loss-of-function alleles linked to brown hilum color. Although our analyses did
detect strong signals in the vicinity of this gene, the most associated markers
were not located in the body of the gene (Figure
\ref{hilum_color_blackbrown_R_signal}). Interestingly, the most significantly
associated \emph{k}-mers detected by our analysis mapped 2 kb upstream of
another gene (Glyma.09g234900) annotated as a MYB transcription factor and
putatively involved in the regulation of anthocyanin biosynthesis. This gene
as well as Glyma.09g235100 and two others are part of a tandem array of four
R2R3 MYB genes considered as candidates by \cite{gillman2011}, and of which
Glyma.09g235100 was the only one expressed in the seed coat. Given the
convincing evidence provided by \cite{gillman2011}, it is most likely that
Glyma.09g235100 is indeed the gene associated with the \emph{R} locus and that
our failure to identify the most significantly associated \emph{k}-mers as
causal variants is due to the coexistence of several variants that can cause
loss of function at the \emph{R} locus. Indeed, we observed at least two causal
variants (a 1-bp deletion and a SNP) previously documented by
\cite{gillman2011} in our dataset (Figure \ref{hilum_color_blackbrown_R_gene},
\ref{hilum_color_blackbrown_R_kmers}).

We conducted a third GWAS focusing on the brown and reddish brown hilum color
phenotypes in order to detect a signal at the \emph{O} locus, however this GWAS
did not detect associations at any of the classical loci (Figure
\ref{hilum_color_rbr_manhattan}).

\subsection*{Pod color}
\label{sv-gwas-pod-color}

We conducted two different GWAS analyses on pod color. The first analysis included four
different colors (Table \ref{POD.COLOR_gwas_table}) and resulted in the
detection of signals overlapping both the \emph{L1} and \emph{L2} loci using all
four methods (Figure \ref{pod_color_all_manhattan}). Detailed results at the
\textit{L1} locus can be viewed in figure \ref{pod_color_all_L1_signal},
whereas detailed results at the \textit{L2} locus can be viewed in figure
\ref{pod_color_all_L2_signal}.

The second GWAS analysis compared black and brown pods with the objective of detecting stronger
signals at the \emph{L1} locus (Table
\ref{POD.COLOR_gwas_table}).  Accordingly, all four methods detected a signal
at the \emph{L1} locus but not at the \emph{L2} locus using this GWAS (Figure
\ref{pod_color_blbr_manhattan}).  Results at the \emph{L1} locus were similar
to those observed when using all observed phenotypes, but the \emph{k}-mer
analysis revealed a previously undocumented signal on chromosome Gm15 which was
not in LD with any other signal (Figures \ref{pod_color_blbr_manhattan},
\ref{pod_color_blbr_ld}). The observed signal at this locus overlaps a total of
27 genes (Figure \ref{pod_color_blbr_pdcGm15_signal}), but none of those are
annotated as having an obvious role in controlling pod color.

\subsection*{Pubescence form}
\label{sv-gwas-pubesence-form}

We conducted two different GWAS analyses on pubescence form. The first analysis used three
phenotypes (erect, semi-appressed and appressed; Table
\ref{PUBESCENCE.FORM_gwas_table}) and detected signals overlapping the
\emph{Pa1} locus using all four approaches as well as signals overlapping the
\emph{Pa2} locus using Platypus and the \emph{k}-mer approach (Figure
\ref{pubescence_form_all_manhattan}). Detailed results at the \textit{Pa1} locus
can be viewed in figure \ref{pubescence_form_all_Pa1_signal},
\ref{pubescence_form_all_Pa1_gene} and \ref{pubescence_form_all_Pa1_kmers},
whereas results at the \textit{Pa2} locus can be viewed in figure
\ref{pubescence_form_all_Pa2_signal}.

Signals were also observed on chromosomes Gm04 and Gm15 by the Platypus and
\emph{k}-mer analyses (Figure \ref{pubescence_form_all_manhattan}), however
upon further analysis of LD patterns they appeared to be linked to the
\emph{Pa1} locus (Figure \ref{pubescence_form_all_ld}).

The second GWAS analysis focused only on appressed and semi-appressed phenotypes
with the objective of enhancing the signal at the \emph{Pa2} locus. However,
this analysis only detected weak signals at the \emph{Pa1} locus (Figure
\ref{pubescence_form_noerect_manhattan}) and its results were therefore not
analyzed further.

\subsection*{Pubescence density}
\label{sv-gwas-pubescence-density}

We conducted a GWAS analysis on pubescence density (Table
\ref{PUBESCENCE.DENSITY_gwas_table}), which detected signals at the \emph{Ps}
locus but not the \emph{Pd1} and \emph{P1} loci using all four approaches
(Figure \ref{pubescence_density_manhattan}). \cite{liu2020ps} linked the gene
Glyma.12g187200 to the \emph{Ps} locus and identified ia 25.6-kb copy number variation
containing this gene as the causal variant, with
higher copy number resulting in decreased pubescence. By
analyzing the sequencing depth in accessions with increased copy number, we
identified the boundaries of the CNV region as 36,254,989 -
36,297,368 (42 kb) on chromosome Gm12.
Detailed results at this locus can be viewed in figures \ref{pubescence-density-main-figure},
\ref{pubescence_density_Ps_signal} and \ref{pubescence_density_Ps_gene}.

\subsection*{Seed coat luster}
\label{sv-gwas-seed-coat-luster}

We performed three different GWAS analyses on seed coat luster (Table
\ref{SEED.COAT.LUSTER_gwas_table}). Since all three analyses yielded similar results
(Figure \ref{seed_coat_luster_all_manhattan},
\ref{seed_coat_luster_nointermediate_manhattan},
\ref{seed_coat_luster_dullshiny_manhattan}), we chose to focus on the
GWAS contrasting the dull and shiny phenotypes as it yielded the most
significant signals. In addition, it also provided results that were most consistent with the
analysis previously done by \cite{bandillo2017}. The analysis detected
signals at the \emph{B} locus as the sole classical locus using all four
methods (Figure \ref{seed_coat_luster_dullshiny_manhattan}). Apparent signals
on chromosomes Gm20 and Gm09 (which may correspond to the B? locus reported by
\cite{bandillo2017}) were also detected, but analysis of the LD patterns
between \emph{k}-mers suggested that these signals were most likely in LD with
the \emph{B} locus (Figure \ref{seed_coat_luster_dullshiny_ld}). All methods
also detected signals overlapping the \emph{Hps} locus as defined by
\cite{bandillo2017}, however results shown by \cite{gijzen2003} and our results
suggest that \emph{Hps} and \emph{B} are the same locus. Using a sequencing depth-based segmentation analysis
similar to the one used for the \emph{Ps} locus, we identified a 31-kb region
spanning positions 9,386,109-9,417,431 on chromosome Gm15 that exhibited
variation in copy number across accessions and overlapped four genes annotated
as containing HPS domains (Figures \ref{seed_coat_luster_dullshiny_B_signal},
\ref{seed_coat_luster_dullshiny_B_gene}). The top 5\% / 1\% significant
associations of all approaches overlapped this CNV region and the
most significantly associated \emph{k}-mer was located within the CNV region
(Figures \ref{seed_coat_luster_dullshiny_B_signal},
\ref{seed_coat_luster_dullshiny_B_gene}). However we were not able to link any
of the variants found to the presence of copy number variation.

% Seed coat luster - I locus

@@

\subsection*{Maturity group}
\label{sv-gwas-maturity-group}

We conducted on a GWAS analysis on maturity group which did not yield any significant signals
overlapping the classical E1-E4 loci using any of the four GWAS approaches.
However, signals at the E1 locus did approach significance in the Platypus
analysis (Table \ref{MATURITY.GROUP_gwas_table}, Figure \ref{maturity_group_manhattan}).

\subsection*{Resistance to \emph{Phytophthora sojae}}
\label{sv-gwas-resistance}

We conducted a GWAS analysis on horizontal resistance to
\textit{Phytophthora sojae} using corrected dry weight (CDW) as a quantitative
measure of resistance. Detailed results can be viewed in figures
\ref{corrected_dry_weight_manhattan} and
\ref{corrected_dry_weight_cdwGm15_signal}.

\subsection*{Seed oil content}
\label{sv-gwas-seed-oil-content}

@@

\subsection*{Seed protein content}
\label{sv-gwas-seed-protein-content}

@@

>>>>

% Detailed results that are not meant to fit the main paper
\section*{Supplemental results}
\label{annexe-sv-gwas-detailed-results}

This section describes results at additional loci that were not discussed in
the main text. A summary of the signals found at all loci using all four
methods is presented in Table \ref{signals_table}.

\subsection*{Flower color -- \textit{L1} locus}
\label{annexe-sv-gwas-flower-color-l1}

In addition to the \textit{W1} locus, \cite{bandillo2017} reported a signal for
flower color near the \textit{L1} locus.  None of the GWAS approaches that we
tested found a signal for flower color near this locus (Figure
\ref{flower_color_manhattan}). This result is unsurprising as W1 is the major
contributor to flower color in soybean, whereas the \textit{L1} locus
(typically associated with pod color) was detected with a much less significant
\textit{p}-value (see Table \ref{signals_table}) and only conjectured as
associated with flower color by \cite{bandillo2017}.

\subsection*{Pubescence color -- locus on chromosome Gm16}
\label{annexe-sv-gwas-pubescence-color-gm16}

In addition to the \textit{T} locus, the first GWAS analysis conducted on pubescence color
revealed a second locus on chromosome Gm16
(Figure \ref{pubescence_color_all_manhattan}). Upon closer analysis, this locus
appeared to be in strong LD with \emph{k}-mers mapping to chromosome Gm01 and
most importantly in moderate LD with the \emph{T} locus (Figure
\ref{pubescence_color_all_ld}), which suggests that this does not represent an
independent locus controlling pubescence color. This interpretation is also
supported by the fact that the signal at that locus only spans about 10 kb on
chromosome Gm16.

\subsection*{Seed coat color -- loci \textit{T}, \textit{O} and \textit{R}}
\label{annexe-sv-gwas-seed-coat-color-t-o}

In addition to the \textit{I} and \textit{G} loci, the \textit{k}-mer-based
GWAS analysis of five different seed coat color phenotypes (first GWAS in Table
\ref{SEED.COAT.COLOR_gwas_table}) detected weak signals overlapping the
\emph{T} and \emph{O} loci (Figure \ref{seed_coat_color_all_manhattan}c).
However, we interpreted those results as chance associations given the low
\textit{p}-value associated with the \emph{T} locus and the proximity of the
\emph{O} locus to the \emph{I} locus. None of the three GWAS approaches detected
a signal overlapping the \textit{R} locus.

\subsection*{Stem termination type -- loci \textit{Dt2} and \textit{E3}}
\label{sv-gwas-stem-termination}

We performed a GWAS analysis comparing only semi-determinate and indeterminate
stem termination phenotypes in an attempt to detect signals at the \emph{Dt2}
locus (Table \ref{STEM.TERMINATION.TYPE_gwas_table}).  However, this was
unsuccessful using all three GWAS approaches (Figure
\ref{stem_termination_sn_manhattan}). Similarly, we failed to detect signals at
the \textit{E3} locus, which is more commonly associated with maturity but was
mentioned as associated with stem termination type by \cite{bandillo2017}.

\subsection*{Hilum color -- loci \textit{W1} and \textit{O}}
\label{sv-gwas-hilum-color}

None of the three GWAS analyses that we conducted on hilum color found signals
at the \emph{W1} and \emph{O} loci studied by \cite{bandillo2017} (Figures
\ref{hilum_color_all_manhattan}, \ref{hilum_color_blackbrown_manhattan},
\ref{hilum_color_rbr_manhattan}).

\subsection*{Pubescence density -- loci \textit{Pd1} and \textit{P1}}
\label{sv-gwas-pubescence-density}

Our GWAS analysis on pubescence density did not detect any signals at the
\emph{Pd1} and \emph{P1} (Figure \ref{pubescence_density_manhattan}). Our
failure to detect these loci is likely due to our small population size
compared to \cite{bandillo2017} and to the smaller contribution of those loci
to pubescence density in our population.

\subsection*{Seed coat luster -- loci \textit{Hps} \textit{B?} and \textit{I}}
\label{sv-gwas-seed-coat-luster}

We performed three different GWAS analyses on seed coat luster (Table
\ref{SEED.COAT.LUSTER_gwas_table}). Since all three analyses yielded similar results
(Figure \ref{seed_coat_luster_all_manhattan},
\ref{seed_coat_luster_nointermediate_manhattan} and
\ref{seed_coat_luster_dullshiny_manhattan}), we chose to focus on the
GWAS contrasting the dull and shiny phenotypes as it yielded the most
significant signals. In addition, it also provided results that were most consistent with the
analysis previously done by \cite{bandillo2017}.

The analysis detected signals at the \emph{B} locus as the sole classical locus
using all four methods (see main text and Figure
\ref{seed_coat_luster_dullshiny_manhattan}).  All methods also detected signals
overlapping the \emph{Hps} locus as defined by \cite{bandillo2017}, however
results shown by \cite{gijzen2003} and our results suggest that \emph{Hps} and
\emph{B} are the same locus.

Apparent signals on chromosomes Gm20 and Gm09 (which may correspond to the B?
locus reported by \cite{bandillo2017}) were also detected, but analysis of the
LD patterns between \emph{k}-mers suggested that these signals are most likely
in LD with the \emph{B} locus (Figure \ref{seed_coat_luster_dullshiny_ld}).

@@

\subsubsection*{Maturity group}
\label{annexe-sv-gwas-maturity}

We did not detect any signals associated with maturity using our dataset
(Figure \ref{maturity_group_manhattan}). The use of maturity group as a proxy
for time to maturity instead of the number of days, combined with the limited
phenotypic variation of this trait in this collection of accessions (Table
\ref{MATURITY.GROUP_gwas_table}), likely explains our failure to detect
any signals.

\subsection*{Seed oil and protein content}
\label{sv-gwas-seed-oil-protein-content}

@@

@@

Our failure to detect signals overlapping the known genes at these two loci for
oil and protein is probably due to the fact that the oil and protein content
data that we used for GWAS were not obtained from orthogonal trials, which was
a clear limitation to this GWAS given our relatively small sample size.

\subsection*{Pubescence form -- additional loci}
\label{sv-gwas-pubesence-form}

The first GWAS analysis on pubescence form found were signals on chromosomes Gm04 and Gm15 with Platypus
\emph{k}-merss (Figure \ref{pubescence_form_all_manhattan}). However,
upon further analysis of LD patterns, they appeared to be linked to the
\emph{Pa1} locus (Figure \ref{pubescence_form_all_ld}).
